---
title: "Comparative analysis for Cox PH model with frailty (gaussian distribution)
  and without fraility on Imputed (MICE and MissForest) and non-imputed datasets."
author: "Sarah Ogutu"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# A plot for proportion of missing variables
```{r}
# setting working directory
setwd( "D:/Imputation")
library(naniar)
library(ggplot2)
library(tidyverse)
# selecting columns with missing values
columns_with_na <- Base_surv_analysis16[, sapply(Base_surv_analysis16,
                                                     function(x) any(is.na(x)))]
# renaming the column names
colnames(columns_with_na) <- c("Amount of income", 
                               "Stable Sex Partners In 30 Days", 
                               "Casual Sex Partners In 30 Days", 
                               "Vaginal Sex", "Anal Sex", "Oral Sex", 
                               "Vaginal & Anal Sex", "Vaginal & Oral Sex", 
                               "New partners", "Age of Oldest Partner", 
                               "Other Sex Partners", "Partner's HIV test", 
                               "Condom use by Stable Partner", 
                               "Condom use by Casual Partner", "HSV2", 
                               "Pain During Sex")
# a plot for prportion of missing values
missing_plot <- gg_miss_var(columns_with_na, show_pct = TRUE)

# saving the plot with high resolution
ggsave("missing_plot.pdf", plot = missing_plot, width = 7, height = 4, 
       units = "in", dpi = 600, device = "pdf")
```

# Stepwise Cox Regression
## Complete Case analysis
```{r}
library(survival)
library(StepReg)
library(kableExtra)
```

### Without Frailty
```{r}
# Complete dataset with NA removed 
Surv_list9.1<- list(Base_surv_analysis16.1.1, df_final1)
Base_surv_analysis23_2<- Surv_list9.1 %>% reduce(inner_join, by='PID') %>%
  na.omit()
Base_surv_analysis23.2 <- subset(Base_surv_analysis23_2,
                             select = -c(PID, months, HIV))
# Define the survival model formula for time-dependent Cox analysis
form3 = Surv(tstart, tstop, event) ~ .

# Perform stepwise variable selection
Time_Cyt_base_results.2 <- stepwise(formula=form3, 
                          data = Base_surv_analysis23.2,
                          type = "cox",
                          strategy = "bidirection",
                          metric =  "AIC",
                          sle = 0.15,
                          sls = 0.15)
# Extract selected variables from stepwise selection
selected_vars <- Time_Cyt_base_results.2$Vote_model$model

# Construct the formula dynamically
form_selected <- as.formula(paste(selected_vars, collapse = " + "))

# Fit the Cox model using the selected variables
ph4 <- coxph(form_selected, data = Base_surv_analysis23.2)

# Table of variables that meet cox PH assumptions
ph4_table1 <- cox.zph(ph4)
ph4_table1.1 <- as.data.frame(ph4_table1$table)
ph4_table1.2 <- ph4_table1.1[!(ph4_table1.1$p>0.05),]
ph4_table1.2 # all met the PH assumption

# Extract AIC value from the model
extractAIC(ph4)

# Exporting to latex
comb4 <- as.data.frame(cbind(
  summary(ph4)$coefficient, summary(ph4)$conf.int))
names(comb4) <- c("coef", "HR", "SE", "Z", 
                       "p_value", "SEE","SEEE", "lower", "upper")
comb4.1 <- subset( comb4, select= c(HR, SE, lower, upper, p_value))
comb4.1 %>% kbl(caption = "Cox Results1", format = "latex", digits = 3)
```

### With Frailty (gaussian distribution)
```{r}
library(coxme)
# Construct the formula dynamically, adding the frailty term
form_frailty <- as.formula(paste(c(selected_vars, "Treat"), 
                                       collapse = " + "),"+ (1 | PID)")

# Fit the Cox model with frailty term
cox_frailty <- coxme(form_frailty, data = Base_surv_analysis23_2)
# Obtain the summary of the fitted Cox frailty model
summary_cox_frail <- summary(cox_frailty)
# Extract the coefficients table from the model summary
coef_table <- summary_cox_frail$coefficients
# Create a dataframe to store Cox model results,
cox_results <- data.frame(
  Variable = rownames(coef_table),
  HR = coef_table[, "exp(coef)"],
  SE = coef_table[, "se(coef)"],
  P_value = coef_table[, "p"]
)
# Compute 95% Confidence Interval for HR
cox_results$`95% CI Lower` <- cox_results$HR * exp(-1.96 * cox_results$SE)
cox_results$`95% CI Upper` <- cox_results$HR * exp(1.96 * cox_results$SE)

rownames(cox_results) <- NULL
# Print results
print(cox_results)
# Reorder columns
cox_results <- cox_results[, c("Variable", "HR", "SE", 
                               "95% CI Lower", "95% CI Upper", "P_value")]
# Export to latex
latex1 <- cox_results %>% kbl(caption = "Fralty1", format = "latex", digits = 3)

```

```{r}
# Extract log-likelihood from the coxme model
logLik_value <- logLik(cox_frailty)

# The number of fixed-effect parameters + 1 for the random effect (frailty)
num_params <- length(fixef(cox_frailty)) + 1  

# Compute AIC
AIC_coxme <- -2 * as.numeric(logLik_value) + 2 * num_params

# Print the AIC value
print(AIC_coxme)
```

## Imputation
### MICE
```{r}
library(mice)
# Create a list containing two datasets
Cyts_base_list <- list( Cytokines, Baselines) 
# Merge the datasets
Cyt_base_missing <- Cyts_base_list %>% reduce(inner_join, by='PID')
Cyt_base_missing1 <- subset(Cyt_base_missing, select = -c(PID))
# Multiple imputation
set.seed(300)
Cyt_base_imputed <- mice(Cyt_base_missing1, m = 5, maxit = 100)
# Extract each of the 5 imputed datasets separately
dataset1 <- complete(Cyt_base_imputed, 1)
dataset2 <- complete(Cyt_base_imputed, 2)
dataset3 <- complete(Cyt_base_imputed, 3)
dataset4 <- complete(Cyt_base_imputed, 4)
dataset5 <- complete(Cyt_base_imputed, 5)
```

### Without Frailty
```{r}
### Results from first imputed dataset
set.seed(0726)
dataset1.1 <-subset(dataset1, select = -c(PID, HIV, months))
Results_data1 <- stepwise(formula = form3,
                                        data =  dataset1.1,
                                        type = "cox",
                                        strategy = "bidirection",
                                        metric =  "AIC",
                                        sle = 0.15,
                                        sls = 0.15)
#Extract selected variables dynamically
selected_vars1 <- Results_data1$Voted_model$model

# Construct Cox model formula dynamically
form_selected1 <- as.formula(paste(selected_vars1, collapse = " + "))

# Fit Cox model using selected variables
res1 <- coxph(form_selected1, data = dataset1.1)

# AIC for this model
extractAIC(res1)
```

```{r}
### Results from second imputed dataset
set.seed(0726)
dataset2.1 <-subset(dataset2, select = -c(PID, HIV, months))
Results_data2 <- stepwise(formula = form3,
                          data =  dataset2.1,
                          type = "cox",
                          strategy = "bidirection",
                          metric =  "AIC",
                          sle = 0.15,
                          sls = 0.15)
#Extract selected variables dynamically
selected_vars2 <- Results_data2$Vote_model$model

# Construct Cox model formula dynamically
form_selected2 <- as.formula(paste(selected_vars2, collapse = " + "))

# Fit Cox model using selected variables
res2 <- coxph(form_selected2, data = dataset2.1)

# AIC for this model
extractAIC(res2)
```

```{r}
### Results from the third dataset
set.seed(0726)
dataset3.1 <-subset(dataset3, select = -c(PID, HIV, months))
Results_data3 <- stepwise(formula = form3,
                          data =  dataset3.1,
                          type = "cox",
                          strategy = "bidirection",
                          metric =  "AIC",
                          sle = 0.15,
                          sls = 0.15)
#Extract selected variables dynamically
selected_vars3 <- Results_data3$Vote_model$model

# Construct Cox model formula dynamically
form_selected3 <- as.formula(paste(selected_vars3, collapse = " + "))

# Fit Cox model using selected variables
res3 <- coxph(form_selected3, data = dataset3.1)

# AIC for this model
extractAIC(res3)
```

```{r}
### Results from the fourth dataset
set.seed(0726)
dataset4.1 <-subset(dataset4, select = -c(PID, HIV, months))
Results_data4 <- stepwise(formula = form3,
                          data =  dataset4.1,
                          type = "cox",
                          strategy = "bidirection",
                          metric =  "AIC",
                          sle = 0.15,
                          sls = 0.15)
#Extract selected variables dynamically
selected_vars4 <- Results_data4$Vote_model$model

# Construct Cox model formula dynamically
form_selected4 <- as.formula( paste(selected_vars4, collapse = " + "))

# Fit Cox model using selected variables
res4 <- coxph(form_selected4, data = dataset4.1)

# AIC for this model
extractAIC(res4)
```

```{r}
### results from the fifth dataset
set.seed(0727)
dataset5.1 <-subset(dataset5, select = -c(PID, HIV, months))
Results_data5 <- stepwise(formula = form3,
                          data =  dataset5.1,
                          type = "cox",
                          strategy = "bidirection",
                          metric =  "AIC",
                          sle = 0.15,
                          sls = 0.15)
#Extract selected variables dynamically
selected_vars5 <- Results_data5$Vote_model$model

# Construct Cox model formula dynamically
form_selected5 <- as.formula(paste(selected_vars5, collapse = " + "))

# Fit Cox model using selected variables
res5 <- coxph(form_selected5, data = dataset5.1)

# AIC for this model
extractAIC(res5)
```

```{r}
### Pooled results
list_res1 <- list(res1, res2, res3, res4, res5)
pooled_results <- pool(as.mira(list_res1))
pooled_summary <- summary(pooled_results)
# Compute HR (exponentiate estimate) and 95% CI
results_df <- pooled_summary %>%
  mutate(
    HR = exp(estimate),                            # Hazard Ratio (HR)
    CI_lower = exp(estimate - 1.96 * std.error),   # 95% CI Lower Bound
    CI_upper = exp(estimate + 1.96 * std.error)    # 95% CI Upper Bound
  ) %>%
  select(term, HR, std.error, CI_lower, CI_upper, p.value) %>%
  rename(SE = std.error)  # Rename standard error column

# View the results dataframe
print(results_df)
# Export to latex
results_df %>% kbl(caption = "Cox Results2", format = "latex", digits = 3)
```

```{r}
# AIC for pooled results
aic_values <- sapply(list_res1, AIC)

# Compute the pooled AIC (average AIC across models)
pooled_AIC <- mean(aic_values)

# Print pooled AIC
print(pooled_AIC)
```

### With Frailty (Gausian distribution)
```{r}
### Results from the first dataset
dataset1.1_F <- dataset1
dataset1.1_F$PID <- as.factor(dataset1.1_F$PID)
# Frailty formula
form_frailty1 <- as.formula(paste(selected_vars1, collapse = " + "), 
                                 "+ (1 | PID)")

# Fit Cox mixed-effects model with frailty term
res1_F <- coxme(form_frailty1, data = dataset1.1_F)
# Extract results into a Data frame
cox_res_F1 <- summary(res1_F)
coef_tab1 <- cox_res_F1$coefficients

res1_F1 <- data.frame(
  Variable = rownames(coef_tab1),
  Est = coef_tab1[, "coef"],
  SE = coef_tab1[, "se(coef)"]
)

rownames(res1_F1) <- NULL
```

```{r}
### Results from second dataset
dataset2.1_Frailty <- dataset2
dataset2.1_Frailty$PID <- as.factor(dataset2.1_Frailty$PID)
# Frailty formula
form_frailty2 <- as.formula(paste(selected_vars2, collapse = " + "), 
                                 "+ (1 | PID)")

# Fit Cox mixed-effects model with frailty term
res2_F <- coxme(form_frailty2, data = dataset2.1_F)
# Extract results into a dataframe
cox_res_F2 <- summary(res2_F)
coef_tab2 <- cox_res_F2$coefficients

res2_F1 <- data.frame(
  Variable = rownames(coef_tab2),
  Est = coef_tab2[, "coef"],
  SE = coef_tab2[, "se(coef)"]
)

rownames(res2_F1) <- NULL
```

```{r}
### Results from the third dataset
dataset3.1_Frailty <- dataset3
dataset3.1_Frailty$PID <- as.factor(dataset3.1_Frailty$PID)
# Frailty formula
form_frailty3 <- as.formula(paste(selected_vars3, collapse = " + "), 
                                 "+ (1 | PID)")

# Fit Cox mixed-effects model with frailty term
res3_F <- coxme(form_frailty3, data = dataset3.1_F)
# Extract results into a dataframe
cox_res_F3 <- summary(res3_F)
coef_tab3 <- cox_res_F3$coefficients

res3_F1 <- data.frame(
  Variable = rownames(coef_tab3),
  Est = coef_tab3[, "coef"],
  SE = coef_tab3[, "se(coef)"]
)

rownames(res3_F1) <- NULL
```

```{r}
### Results from the fourth dataset
dataset4.1_Frailty <- dataset4
dataset4.1_Frailty$PID <- as.factor(dataset4.1_Frailty$PID)
# Frailty formula
form_frailty4 <- as.formula(paste(selected_vars4, collapse = " + "), 
                                 "+ (1 | PID)")

# Fit Cox mixed-effects model with frailty term
res4_F <- coxme(form_frailty4, data = dataset4.1_F)
# Extract results into a dataframe
cox_res_F4 <- summary(res4_F)
coef_tab4 <- cox_res_F4$coefficients

res4_F1 <- data.frame(
  Variable = rownames(coef_tab4),
  Est = coef_tab4[, "coef"],
  SE = coef_tab4[, "se(coef)"]
)

rownames(res4_F1) <- NULL
```

```{r}
### Results from the fifth dataset
dataset5.1_Frailty <- dataset5
dataset5.1_Frailty$PID <- as.factor(dataset5.1_Frailty$PID)
# Frailty formula
form_frailty5 <- as.formula(paste(selected_vars5, collapse = " + "), 
                                 "+ (1 | PID)")

# Fit Cox mixed-effects model with frailty term
res5_F <- coxme(form_frailty5, data = dataset5.1_F)
# Extract results into a dataframe
cox_res_F5 <- summary(res5_F)
coef_tab5 <- cox_res_F5$coefficients
res5_F1 <- data.frame(
  Variable = rownames(coef_tab5),
  Est = coef_tab5[, "coef"],
  SE = coef_tab5[, "se(coef)"]
)

rownames(res5_F1) <- NULL
```

```{r}
### Pooled results for frailty 
# list the dataframes
res_list <- list(res1_F1, res2_F1, res3_F1, res4_F1, res5_F1)
# Combine the data into a single data frame
combined <- do.call(rbind, res_list)

# Compute pooled estimates using Rubin's Rules
pooled <- combined %>%
  group_by(Variable) %>%
  summarise(
    pooled_coef = mean(Est),
    HR = exp(pooled_coef),
    pooled_se = sqrt(mean(SE^2) + var(Est)),  # Rubin's rule for combining SEs
    .groups = 'drop'
  )

pooled <- pooled %>%
  mutate(
    # 95% Confidence Intervals
    lower_ci = exp(pooled_coef - 1.96 * pooled_se),
    upper_ci = exp(pooled_coef + 1.96 * pooled_se),
    
    # z-score for p-value
    z_value = pooled_coef / pooled_se,
    
    # Two-tailed p-value from the normal distribution
    p_value = 2 * (1 - pnorm(abs(z_value)))
  )

# View pooled coefficients, CIs, and p-values
pooled
# Select the relevant columns
pooled1 <- pooled %>% select(Variable, HR, pooled_se, lower_ci, upper_ci, p_value)
# Export to latex
pooled1 %>% kbl(caption = "model4", format = "latex", digits = 3)

```

```{r}
### Pooled AIC for frailty models
coxme_models <- list(res1_F, res2_F, res3_F, res4_F, res5_F)

# Compute AIC for each model
aic_values <- sapply(coxme_models, function(model) {
  logLik_value <- logLik(model)
  num_params <- length(fixef(model)) + 1  # Fixed + random effect parameters
  -2 * as.numeric(logLik_value) + 2 * num_params  # AIC formula
})

# Compute pooled AIC (average AIC)
pooled_AIC <- mean(aic_values)

# Print pooled AIC
print(pooled_AIC)

```

### MissForest
```{r}
library(missForest)
# Impute
timedp_complete <- missForest(Cyt_base_missing, 
           maxiter = 100, ntree = 100, 
           verbose = TRUE)
timedp_complete_forest <- Baseline_var_imputed_miss$ximp

```

### Without Frailty (Gaussian distribution)
```{r}
set.seed(2409)
timedp_complete_forest1 <- subset(timedp_complete_forest, select = -c(PID))
# Stepwise Cox regression
timedp_base_imputed_forest <- stepwise(formula=form3, 
                                       data =  timedp_complete_forest1,
                                       type = "cox",
                                       strategy = "bidirection",
                                       metric =  "AIC",
                                       sle = 0.15,
                                       sls = 0.15)

# Extract selected variables from stepwise selection
selected_vars6 <- timedp_base_imputed_forest$Vote_model$model

# Create a formula for the Cox PH model
form_selected6 <- as.formula(paste(selected_vars6, collapse = " + "))

# Fit the Cox PH model using the formula
ph_M4 <- coxph(form_selected6, data = timedp_complete_forest)

# Table of variables that meet cox PH assumptions
ph_M4_table1 <- cox.zph(ph_M4)
ph_M4_table1.1 <- as.data.frame(ph_M4_table1$table)
ph_M4_table1.2 <- ph_M4_table1.1[!(ph_M4_table1.1$p>0.05),]
ph_M4_table1.2 # all met the PH assumption

# Extract AIC value from model 1
extractAIC(ph_M4)

# Exporting to latex
comb_M4 <- as.data.frame(cbind(
  summary(ph_M4)$coefficient, summary(ph_M4)$conf.int))
names(comb_M4) <- c("coef", "HR", "SE", "Z", 
                       "p_value", "SEE","SEEE", "lower", "upper")
comb_M4_1.2 <- subset( comb_M4, select= c(HR, SE, lower, upper, p_value))
comb_M4_1.2 %>% kbl(caption = "model1C", format = "latex", digits = 3)

```

### With frailty (Gaussian distribution)
```{r}
data4 <- timedp_complete_forest2
data4$PID <- as.factor(data4$PID)

# Frailty formula
form_frailty6 <- as.formula(paste(selected_vars6, collapse = " + "),
                            "+ (1 | PID)")
# Cox frailty model
cox_frailty4 <- coxme(form_frailty6, data = timedp_complete_forest)

# Data frame of results
summary_cox_frail2 <- summary(cox_frailty4)
coef_table2 <- summary_cox_frail2$coefficients
cox_results2 <- data.frame(
  Variable = rownames(coef_table2),
  HR = coef_table2[, "exp(coef)"],
  SE = coef_table2[, "se(coef)"],
  P_value = coef_table2[, "p"]
)
# Compute 95% Confidence Interval for HR
cox_results2$`95% CI Lower` <- cox_results2$HR * exp(-1.96 * cox_results2$SE)
cox_results2$`95% CI Upper` <- cox_results2$HR * exp(1.96 * cox_results2$SE)

rownames(cox_results2) <- NULL
# Print results
print(cox_results2)
# Export to latex
cox_results2 <- cox_results2[, c("Variable", "HR", "SE", 
                                 "95% CI Lower", "95% CI Upper", "P_value")]
latex3 <- cox_results2 %>% kbl(caption = "Fralty3", format = "latex", digits = 3)

```

```{r}
### AIC for the frailty model from MissForest imputed dataset
# Extract log-likelihood from the coxme model
logLik_value2 <- logLik(cox_frailty4)

# Count the number of fixed-effect parameters + 1 for the random effect (frailty)
num_params2 <- length(fixef(cox_frailty4)) + 1  

# Compute AIC
AIC_coxme2 <- -2 * as.numeric(logLik_value2) + 2 * num_params2

# Print the AIC value
print(AIC_coxme2)
```

## Comparing the two imputation methods
```{r}
# train and test dataset
df_final2 <- subset(df_final1, select = -c(IP_10))
Base_miss_complete_train = df_final2
Base_miss_complete_train$HIV <- as.factor(Base_miss_complete_train$HIV)

set.seed(2409)
train_Index <- createDataPartition(Base_miss_complete_train$HIV,p=.70,
                                  list=FALSE)
train <- Base_miss_complete_train[trainIndex,]
test  <- Base_miss_complete_train[-trainIndex,]

train_X <- subset(train, select = -c(HIV))
test_X <- subset(test, select = -c(HIV))
```

### MissForest Trainning and testing
```{r}
library(randomForest)
# 1) impute train
imp_train_X <- missForest(train_X)$ximp

# 2) build model
rf <- randomForest(x = imp_train_X, y = train$HIV)

# 3) & 4) combine & impute test
train_test_X <- rbind(test_X, imp_train_X)
imp_test_X <- missForest(train_test_X)$ximp[1:nrow(test_X), ]

# 5) predict for test
pred_test <- predict(rf, imp_test_X, type = "prob")

# 6) test ROC & AUC
test_scores <- data.frame(event_prob = pred_test[ ,2], labels = test$HIV)

library(PRROC)
test_roc_v1 <- roc.curve(scores.class0 = 
                           test_scores[test_scores$labels == "1", ]$event_prob,
                         # scores for the POSITIVE class
                      scores.class1 = test_scores[test_scores$labels == "0",
                                                  ]$event_prob, 
                      # scores for the NEGATIVE class
                      curve=T)
test_roc_v1$auc 
```

### Mice training and testing
```{r}
# 1) impute train
imp_train_X2 <- complete(mice(train_X, m=5, 
                             maxit = 100, set.seed(0724)))
imp_test_X2 <- complete(mice(test_X, m=5, 
                             maxit = 100, set.seed(0724)))
# 3) build model
rf2 <- randomForest(x = imp_train_X2, y = train$HIV)

# 4) predict for test
pred_test2 <- predict(rf2, imp_test_X2, type = "prob")

# 5) test ROC & AUC
test_scores2 <- data.frame(event_prob2 = pred_test2[ ,2], labels = test$HIV)

test_roc_v2 <- roc.curve(scores.class0 = 
                           test_scores2[test_scores2$labels == "1",
                                        ]$event_prob2, 
                         # scores for the POSITIVE class
                      scores.class1 = test_scores2[test_scores2$labels == "0",
                                                   ]$event_prob2, 
                      # scores for the NEGATIVE class
                      curve=T)
test_roc_v2$auc
```

### ROC curve 
```{r}
data.frame(method = "missForest", FPR = test_roc_v1$curve[ ,1], TPR = test_roc_v1$curve[ ,2]) %>%
  rbind(data.frame(method = "mice", FPR = test_roc_v2$curve[ ,1], TPR = test_roc_v2$curve[ ,2])) %>%
  ggplot(aes(x = FPR, y = TPR, col = method)) + 
  geom_line() + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") + 
  labs(title = "MissForest vs mice Imputation", 
       subtitle = paste0("MissForest AUC = ", 
                         round(test_roc_v1$auc, 3), ", mice AUC = ",
                         round(test_roc_v2$auc, 3)), 
       col = "Imputation Method")

frame <- data.frame(method = "missForest", FPR = test_roc_v1$curve[ ,1], 
                    TPR = test_roc_v1$curve[ ,2]) %>%
  rbind(data.frame(method = "mice", FPR = test_roc_v2$curve[ ,1], 
                   TPR = test_roc_v2$curve[ ,2])) 

Imp_plot <- frame %>%
  ggplot(aes(x = FPR, y = TPR, col = method)) + 
  geom_line() + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") + 
  labs(col = "Imputation Method")

ggsave("Imp_plot.pdf", plot = Imp_plot, width = 7, height = 4, 
       units = "in", dpi = 600, device = "pdf")
```

